Plant Trait Imputation

### 2024-05-29 code/L1/L1_P5_TropicalAndes_plant_imputetraits.Rmd
rename L1 plant scripts to update number order

## Create a dataframe of species and TraitNames with TraitValue = NA
```{r}
# remove X column
plant_traits <- plant_traits[, !colnames(plant_traits) %in% "X", drop = FALSE]
# Convert the wide dataframe to long format
long_plant_traits <- plant_traits %>%
  gather(TraitName, TraitValue, -species)

@@

## rename trait_value column so it's consistent with other traits
Fruit.Type_families_BIEN_df$trait_value <- Fruit.Type_families_BIEN_df$trait_value_mode_trait
Growth.Form_families_BIEN_df$trait_value <- Growth.Form_families_BIEN_df$trait_value_mode_trait

## convert numeric trait values into character
Plant.Height_families_BIEN_df$trait_value <- as.character(Plant.Height_families_BIEN_df$trait_value)
Fruit.Length_families_BIEN_df$trait_value <- as.character(Fruit.Length_families_BIEN_df$trait_value)

all_family_traits <- bind_rows(
  select(Dispersal.Syndrome_families_BIEN_df, scrubbed_family, trait_name, trait_value),
  select(Fruit.Length_families_BIEN_df, scrubbed_family, trait_name, trait_value),
  select(Plant.Height_families_BIEN_df, scrubbed_family, trait_name, trait_value),
  select(Fruit.Type_families_BIEN_df, scrubbed_family, trait_name, trait_value),
  select(Growth.Form_families_BIEN_df, scrubbed_family, trait_name, trait_value)
  )

@@

## rename trait_value column so it's consistent with other traits
Fruit.Type_genus_BIEN_df$trait_value <- Fruit.Type_genus_BIEN_df$trait_value_mode_trait
Growth.Form_genus_BIEN_df$trait_value <- Growth.Form_genus_BIEN_df$trait_value_mode_trait

## convert numeric trait values into character
Plant.Height_genus_BIEN_df$trait_value <- as.character(Plant.Height_genus_BIEN_df$trait_value)
Fruit.Length_genus_BIEN_df$trait_value <- as.character(Fruit.Length_genus_BIEN_df$trait_value)

all_genus_traits <- bind_rows(
  select(Dispersal.Syndrome_genus_BIEN_df, scrubbed_genus, trait_name, trait_value),
  select(Fruit.Length_genus_BIEN_df, scrubbed_genus, trait_name, trait_value),
  select(Plant.Height_genus_BIEN_df, scrubbed_genus, trait_name, trait_value),
  select(Fruit.Type_genus_BIEN_df, scrubbed_genus, trait_name, trait_value),
  select(Growth.Form_genus_BIEN_df, scrubbed_genus, trait_name, trait_value)
  )

@@

make sure column names match between dataframes before combining
```{r}
colnames(all_family_traits)
colnames(all_genus_traits)
colnames(long_plant_traits_tax)
```

@@

use family & genus level traits to fill in gaps of original dataframe
genus level prefered over family
```{r}
# Join the dataframes based on genus and TraitName
filled_long_plant_traits_tax <- long_plant_traits_tax %>%
  left_join(all_genus_traits, by = c("genus", "TraitName"), relationship = "many-to-many") %>%
  # Fill NAs in TraitValue column with values from all_genus_traits
  mutate(TraitValue = ifelse(is.na(TraitValue.x), TraitValue.y, TraitValue.x)) %>%
  # Remove the columns from all_genus_traits (if needed)
  select(-TraitValue.y)
```

Add units to wide trait table numeric trait names
fruit mass -> Fruit.Mass_mg
fruit length -> Fruit.Length_mm
plant height -> Plant.Height_m
plant lifespan -> Plant.Lifespan_years
seed mass -> Seed.Mass_mg
seed length -> Seed.Length_mm
seed width -> Seed.Width_mm

@@

```{r}
#replace " " with _ in column names
names(plant_traits) <- sub(" ", "_", names(plant_traits))
```

# show the missing data pattern
```{r}
md.pattern(plant_traits)
```
# Trying to prevent error
Error in edit.setup(data, setup, ...) : 
  `mice` detected constant and/or collinear variables. No predictors were left after their removal.
```{r}
# Subset to numeric plant traits
num_plant_traits <- plant_traits %>%
  select_if(is.numeric)
# Check for constant variables
constant_vars <- sapply(num_plant_traits, function(x) length(unique(x))) == 1
constant_vars <- names(constant_vars[constant_vars])

# Remove constant variables
num_plant_traits <- num_plant_traits[, !names(num_plant_traits) %in% constant_vars]

# Check for collinearity
cor_matrix <- cor(num_plant_traits)

# Find highly correlated variables
highly_correlated_pairs <- which(abs(cor_matrix) > 0.8 & upper.tri(cor_matrix, diag = TRUE), arr.ind = TRUE)

# Print highly correlated pairs
print(highly_correlated_pairs)
```

# Perform the imputation
```{r}
imp_model <- mice(plant_traits, method = "cart", maxit = 20)
```

@@

>>>>

## Create a dataframe of species and TraitNames with TraitValue = NA
```{r}
# remove X column
plant_traits <- plant_traits[, !colnames(plant_traits) %in% "X", drop = FALSE]
# remove Growth Forme trait
plant_traits <- plant_traits %>%
  select(-Growth.Forme)
# Convert the wide dataframe to long format
long_plant_traits <- plant_traits %>%
  gather(TraitName, TraitValue, -species)

@@

## rename trait_value column so it's consistent with other traits
Fruit.Type_families_BIEN_df$TraitValue <- Fruit.Type_families_BIEN_df$trait_value_mode_trait
Growth.Form_families_BIEN_df$TraitValue <- Growth.Form_families_BIEN_df$trait_value_mode_trait

## convert numeric trait values into character
Plant.Height_families_BIEN_df$TraitValue <- as.character(Plant.Height_families_BIEN_df$TraitValue)
Fruit.Length_families_BIEN_df$TraitValue <- as.character(Fruit.Length_families_BIEN_df$TraitValue)

all_family_traits <- bind_rows(
  select(Dispersal.Syndrome_families_BIEN_df, scrubbed_family, trait_name, TraitValue),
  select(Fruit.Length_families_BIEN_df, scrubbed_family, trait_name, TraitValue),
  select(Plant.Height_families_BIEN_df, scrubbed_family, trait_name, TraitValue),
  select(Fruit.Type_families_BIEN_df, scrubbed_family, trait_name, TraitValue),
  select(Growth.Form_families_BIEN_df, scrubbed_family, trait_name, TraitValue)
  )

@@

## rename TraitValue column so it's consistent with other traits
Fruit.Type_genus_BIEN_df$TraitValue <- Fruit.Type_genus_BIEN_df$trait_value_mode_trait
Growth.Form_genus_BIEN_df$TraitValue <- Growth.Form_genus_BIEN_df$trait_value_mode_trait

## convert numeric trait values into character
Plant.Height_genus_BIEN_df$TraitValue <- as.character(Plant.Height_genus_BIEN_df$TraitValue)
Fruit.Length_genus_BIEN_df$TraitValue <- as.character(Fruit.Length_genus_BIEN_df$TraitValue)

all_genus_traits <- bind_rows(
  select(Dispersal.Syndrome_genus_BIEN_df, scrubbed_genus, trait_name, TraitValue),
  select(Fruit.Length_genus_BIEN_df, scrubbed_genus, trait_name, TraitValue),
  select(Plant.Height_genus_BIEN_df, scrubbed_genus, trait_name, TraitValue),
  select(Fruit.Type_genus_BIEN_df, scrubbed_genus, trait_name, TraitValue),
  select(Growth.Form_genus_BIEN_df, scrubbed_genus, trait_name, TraitValue)
  )

@@

rename traits to match
```{r}
# Replace trait names to match
long_plant_traits_tax <- long_plant_traits_tax %>%
  mutate(TraitName = ifelse(TraitName == "Plant.Height", "Plant.Height_m", TraitName)) %>%
  mutate(TraitName = ifelse(TraitName == "Fruit.Length", "Fruit.Length_mm", TraitName))
unique(long_plant_traits_tax$TraitName)

na_traits_family.genus <- na_traits_family.genus %>%
  mutate(TraitName = ifelse(TraitName == "Plant.Height", "Plant.Height_m", TraitName)) %>%
  mutate(TraitName = ifelse(TraitName == "Fruit.Length", "Fruit.Length_mm", TraitName))
unique(na_traits_family.genus$TraitName)
```

make sure column names match between dataframes before combining
```{r}
colnames(all_family_traits)
colnames(all_genus_traits)
colnames(long_plant_traits_tax)
colnames(na_traits_family.genus)
```

@@

use family & genus level traits to fill in gaps of na_traits_family.genus
genus level prefered over family
```{r}
# Join the dataframes based on genus and TraitName
filled_na_traits_family.genus <- na_traits_family.genus %>%
  left_join(all_genus_traits, by = c("genus", "TraitName"), relationship = "many-to-many") %>%
  # Fill in TraitValue column with values from all_genus_traits
  mutate(TraitValue = coalesce(TraitValue.x, TraitValue.y)) %>%
  # Remove the columns
  select(-TraitValue.y, -TraitValue.x)
```

```{r}
na_traits_after_genus <- filled_na_traits_family.genus %>%
  filter(is.na(TraitValue))
filled_traits_with_genus <- filled_na_traits_family.genus %>%
  filter(!is.na(TraitValue))
```

```{r}
# Join the dataframes based on family and TraitName
## remove trait level column from na_traits_after_genus
na_traits_after_genus <- na_traits_after_genus %>%
  select(-TraitLevel)
filled_na_traits_family <- na_traits_after_genus %>%
  left_join(all_family_traits, by = c("family", "TraitName"), relationship = "many-to-many") %>%
  # Fill in TraitValue column with values from all_genus_traits
  mutate(TraitValue = coalesce(TraitValue.x, TraitValue.y)) %>%
  # Remove the columns
  select(-TraitValue.y, -TraitValue.x)
```

```{r}
filled_traits_with_family <- filled_na_traits_family %>%
  filter(!is.na(TraitValue))
na_traits_after_family <- filled_na_traits_family %>%
  filter(is.na(TraitValue))
```

```{r}
nonimputed_traits <- long_plant_traits_tax %>%
  filter(!is.na(TraitValue))
```

combine filled_traits_with_genus, filled_traits_with_family, and nonimputed_traits
```{r}
colnames(filled_traits_with_genus)
colnames(filled_traits_with_family)
colnames(nonimputed_traits)
```

```{r}
all_traits_with_tax <- rbind(filled_traits_with_genus, filled_traits_with_family, nonimputed_traits)
```

```{r}
all_traits_with_tax %>%
  group_by(TraitName) %>%
  summarise(num_species = n_distinct(species))
```

```{r}
na_traits_after_family %>%
  group_by(TraitName) %>%
  summarise(num_species = n_distinct(species))
```

add NAs back to all traits
```{r}
colnames(all_traits_with_tax)
colnames(na_traits_after_family)
```

```{r}
all_traits_with_NAs <- rbind(all_traits_with_tax, na_traits_after_family)
```

### Save long dataframe
```{r}
write.csv(all_traits_with_NAs, file.path(output_path,"TropicalAndes_all_plant_traits_filled_with_family_genus_long.csv"))
```

## Convert long dataframe to wide
```{r}
#remove unused columns
all_traits_with_NAs_long <- all_traits_with_NAs %>%
  select(-TraitLevel, -family, -genus)
wide_traits <- pivot_wider(all_traits_with_NAs_long, names_from = TraitName, values_from = TraitValue)
```
make NA if:
- Growth.Form is numeric
- Dispersal.Syndrome is numeric or contains multiple values
- Plant.Height = 0
- Fruit.Type is numeric
```{r}
library(stringr)
wide_traits$Growth.Form <- ifelse(is.numeric(wide_traits$Growth.Form), NA, wide_traits$Growth.Form)
wide_traits$Dispersal.Syndrome[stringr::str_detect(wide_traits$Dispersal.Syndrome, ",")] <- NA

# Convert Plant.Height_m to numeric 
wide_traits$Plant.Height_m[wide_traits$Plant.Height_m == "0"] <- NA

wide_traits$Fruit.Type <- ifelse(is.numeric(wide_traits$Fruit.Type), NA, wide_traits$Fruit.Type)
```

```{r}
wide_traits <- wide_traits %>%
  mutate_all(~sapply(., paste, collapse = ", "))

# Replace values with multiple values with NA
wide_traits <- wide_traits %>%
  mutate(Dispersal.Syndrome = if_else(str_count(Dispersal.Syndrome, ", ") > 0, NA_character_, Dispersal.Syndrome), 
         Growth.Form = if_else(str_count(Growth.Form, ", ") > 0, NA_character_, Growth.Form),
         Plant.Height_m = if_else(str_count(Plant.Height_m, ", ") > 0, NA_character_, Plant.Height_m), 
         Fruit.Length_mm = if_else(str_count(Fruit.Length_mm, ", ") > 0, NA_character_, Fruit.Length_mm), 
         Fruit.Type = if_else(str_count(Fruit.Type, ", ") > 0, NA_character_, Fruit.Type))
```

```{r}
wide_traits$Plant.Height_m <- unlist(wide_traits$Plant.Height_m)
wide_traits$Growth.Form <- unlist(wide_traits$Growth.Form)
wide_traits$Dispersal.Syndrome <- unlist(wide_traits$Dispersal.Syndrome)
wide_traits$Fruit.Length_mm <- unlist(wide_traits$Fruit.Length_mm)
wide_traits$Fruit.Type <- unlist(wide_traits$Fruit.Type)

#convert column types from list to character & numeric
wide_traits$Plant.Height_m <- as.numeric(wide_traits$Plant.Height_m)
wide_traits$Growth.Form <- as.character(wide_traits$Growth.Form)
wide_traits$Dispersal.Syndrome <- as.character(wide_traits$Dispersal.Syndrome)
wide_traits$Fruit.Length_mm <- as.numeric(wide_traits$Fruit.Length_mm)
wide_traits$Fruit.Type <- as.character(wide_traits$Fruit.Type)
```

@@

# show the missing data pattern
```{r}
md.pattern(wide_traits)
```
# Perform the imputation
```{r}
imp_model <- mice(wide_traits, method = "cart", maxit = 20)
```

@@

remove dispersal syndrome because I need to fix it 
```{r}
imputed_data <- imputed_data %>%
  select(-Dispersal.Syndrome)
```

```{r}
imputed_data %>%
  summarise_at(vars(Growth.Form:Fruit.Type), ~ sum(is.na(.)))
```

# Save imputed data
```{r}
write.csv(imputed_data, file.path(output_path,"TropicalAndes_imputed_plant_traits.csv"))
```
